A comparison of single-cell trajectory inference methods: towards more accurate and robust tools

### 2018-02-07 analysis/paper/paper.Rmd

@@

```{r}
add_sfig({
n_methods_over_time <- readRDS(figure_file("n_methods_over_time.rds", experiment_id="method_characteristics"))
trajectory_components_over_time <- readRDS(figure_file("trajectory_components_over_time.rds", experiment_id="method_characteristics"))

@@

<!-- history of Ti, different methods -->
Dozens of trajectory inference methods have been developed over the last years, and more are being published almost every month (`r ref("sfig", "methods_over_time", "a")`). Initially, most trajectory inference fixed the trajectory topology to a linear or bifurcating topology, by either algorithmic limitations or through user parameters, and focussed on correctly ordering the cells along this fixed topology. However, several recent methods now infer both the ordering and the topology at the same time, although certain structural limitations are still imposed(`r ref("sfig", "methods_over_time", "b")`). Methods which do not impose limitations on the topology are now being developed [@welch_slicer:_2016; @wolf_graph_2017], and it is expected that in the future methods will be able to model even more complex behavior, such as multiple dynamic processes happening at parallel in a single-cell or the integration of datasets from different patients [@tanay_scaling_2017; @cannoodt_computational_2016].

<!-- Common framework of TI, different components, naturally progressing towards the need for an evaluation -->
Although trajectory inference methods use a variety of algorithms and subcomponents to reach a final ordering, most consist of the following three steps: (i) preprocessing (normalization, filtering of genes and/or cells), (ii) conversion to a simplified representation using dimensionality reduction, clustering or graph building and (iii) ordering the cells along the simplified representation [@cannoodt_computational_2016]. Depending on the way the cells are ordered, some prior information about the dynamic process can be optionally used or required by the method, which can both bias the trajectory to current knowledge but also guide the method towards choosing the right trajectory among many others. Furthermore, with the ongoing strong scalability in single-cell -omics technologies, it becomes more and more important that analysis methods become scalable and user friendly .

@@

```{r}
# Load method characteristics
methods <- readRDS(derived_file("methods.rds", experiment_id="method_characteristics"))
methods_evaluated <- readRDS(derived_file("methods_evaluated.rds", experiment_id="method_characteristics"))
```

We gathered a list of `r sum(methods$is_ti, na.rm=TRUE)` TI methods, and selected a subset of `r sum(methods$evaluated)` for evaluation, primarily based on their free availability and the presence of a programming interface (`r ref("table", "methods")`). We characterized all methods in four different ways: possible trajectory topologies, implementation quality, prior information and the underlying algorithm (`r ref("fig", "method_characteristics")`).

`r ref("table", "methods", anchor=TRUE)` Overview of current trajectory inference methods, and whether they were included in this current evaluation study
```{r}
read_rds(figure_file("methods_table.rds", "method_characteristics"))
```

At the level of trajectory topologies, most recent methods can handle one or more splits (either bifurcations or multifurcations) in the trajectory (`r ref("fig", "method_characteristics", "a")`). However, only a handful, mostly recent, methods can currently handle more complex graph topologies which can include loops (`r ref("sfig", "methods_over_time", "b")`), one of which is included in the current evaluation (SLICER).

@@

```{r}
add_stable(
read_rds(figure_file("qc_checks.rds","method_characteristics")), 
"qc_checks", 
""
)
```

`r anchor("fig", "method_characteristics")`
![](`r figure_file("method_characteristics.svg", experiment_id="method_characteristics")`)
`r ref("fig", "method_characteristics", anchor=TRUE)` First characterization of each method.

```{r}
add_sfig({
readRDS(figure_file("method_qc_check.rds", experiment_id="method_characteristics"))
}, 
"method_qc_check", 
"Caption", 
10, 
13
)
```

Most methods fulfill most of the basic criteria, such as free availability and basic code quality criteria (`r ref("fig", "method_characteristics", "b")`). Recent methods tend to have higher code quality (`r ref("sfig", "qc_over_time")`). However, several aspects are consistently lacking for the majority of the methods (`r ref("fig", "method_characteristics", "e")`) and we believe that these should receive extra attention from developers. Although these outstanding issues cover all five categories, code assurance and documentation in particular are problematic areas (`r ref("sfig", "method_qc_check")` and `r ref("fig", "method_characteristics", "c")`), notwithstanding several studies pinpointing these as good practices [@wilson_best_2014; @artaza_top10metrics_2016].

```{r}
add_sfig({
readRDS(figure_file("qc_over_time.rds", experiment_id="method_characteristics"))
}, 
"qc_over_time", 
"Quality control scores and number of citations over time", 
10, 
10
)
```

@@

```{r}
add_sfig({
readRDS(figure_file("method_components.rds", experiment_id="method_characteristics"))
}, 
"method_components", 
"TI methods frequently share steps", 
15, 
7.5
)
```
## Evaluating the accuracy of the inferred trajectory

@@

```{r}
add_stable(
read_rds(figure_file("datasets.rds","dataset_characterisation")), 
"datasets", 
"Real datasets used in this study"
)
```

# Discussion
Limitations of the model:

* Robrecht.Cannoodt @robrechtc 10:21
en heterogene output van alle methode... true, maar sommigen hebben we wel wat moeten tweaken 😛
en ik kan me output bedenken dat niet in ons model zou kunnen passen
* Wouter Saelend @wouters 10:21
oh 😛
* Robrecht.Cannoodt @robrechtc 10:21
welja
* Wouter Saelend @wouters 10:21
Misschien een kritiekpunt is onzekerheid... maarja 😛
* Robrecht.Cannoodt @robrechtc 10:22
onzekerheid, en dat een cell in meerdere niet-neighbouring milestones zou kunnen zitten
als in, biologisch gezien houdt het mss geen steek, maar een methode zou wel kunnen zeggen, "wete, ik dénk dat het ofwel een DC is of een rode bloedcel"
* Wouter Saelend @wouters 10:24
hmmm uhu true, hoort allemaal wat samen met de onzekerheid. Je zou per cell een probabiliteitsdistributie doorheen het milestone netwerk kunnen geven, maar sla dat eens op 😛
* Robrecht.Cannoodt @robrechtc 10:25
uhu ^^
* Wouter Saelend @wouters 10:25
En bereken dan eens de emd 😛
* Robrecht.Cannoodt @robrechtc 10:25
idd...
deze discussie zou wel ergens een plaatsje kunnen krijgen in het manuscript, vind ik
* Wouter Saelend @wouters 10:25
Uhu in de discussie dan wss
* Robrecht.Cannoodt @robrechtc 10:26
tiens, een discussie die in de discussie zou plaats vinden, wat een goed idee 😮
<tease>

@@

```{r}
add_sfig({
readRDS(figure_file("trajectory_type_trees.rds", experiment_id="dataset_characterisation/trajectory_types"))
}, 
"trajectory_type_tree", 
"Different trajectory types can be classified in a directed acyclic graph, in which certain types are a generalisation of other trajectory types.", 
15, 
10
)
```

@@

* Methods returning a single pseudotime value: `r conversion("linear")`. A single edge between two milestones, its percentage proportional to the pseudotime.
* Methods which assign each cell to a branch together with a pseudotime value and a branch network: `r conversion("branching_local")`. Similarly, the branch network is used as the milestone network, the percentages of a cell are proportional with its branch pseudotime.
* Methods returning both a global pseudotime value and a probability for every end state: `r conversion("branching_global")`. We use a single start state and an add an edge to every end milestone each representing an end state. The global pseudotime then determines the distance from the begin milestone, the rest of the cell's position is calculated by distributing the residual percentage over the end states, proportionally to the end state probabilities.
* Methods returning a cluster assignment and a cluster network:  `r conversion("branching_cluster")`. The cluster network was used as milestone network, each cell received percentage 1 or 0 based on its cluster assignment.
* Methods returning a cluster assignment, cluster network and dimensionality reduction: `r conversion("branching_local_projection")`. We projected each cell on the closest point on the edges between its own cluster and neighbouring clusters within the dimensionality reduction. This projection allowed us to give each cell a pseudotime within the edge, which was then converted into our model as described above, using the cluster network as milestone network.

@@

* DPT: We projected the cells onto the cluster network, consisting of a central milestone (this cluster contains the cells which were assigned to the "unknown" branch) and three terminal milestones, each corresponding to 
* Sincell: To constrain the number of milestones this method creates, we merged two cell clusters iteratively until the percentage of leaf nodes was below a certain cutoff, default at 25%. This was then processed 
* SLICER: Although this method returns a branch network and a global pseudotime, it was not clear to us how to assign cells to individual branches in the network. We therefore used the cell graph provided by the method to calculate the shortest paths between each branch, and then projected other cells onto this backbone graph structure, creating a local pseudotime for every cell within its branch. The output was then further processed as above.
* SLICE: As discussed in the vignette of SLICE, we ran principal curves one by one for every edge detected by SLICE. This local pseudotime was then processed as above.
* MFA: We used the branch assignment as state probabilities, which together with the global pseudotime were processed as described above.

@@

```{r}
add_sfig({
readRDS(figure_file("modulenets.rds", experiment_id="dataset_characterisation/synthetic"))
}, 
"modulenets", 
"Module networks, milestone networks and the module dynamics for each of the different types of synthetic data", 
15, 
5
)
```

@@

```{r}
add_stable(
read_rds(figure_file("samplers.rds","dataset_characterisation/synthetic")), 
"samplers", 
"Distributions from which each parameter in the thermodynamic model was sampled"
)
```

@@

```{r}
add_sfig({
readRDS(figure_file("score_aspects.rds", experiment_id="score_rules/score_aspects"))
}, 
"score_aspects", 
"Different scores assess different aspects of the ", 
15, 
8
)
```

@@

```{r}
add_sfig({
readRDS(figure_file("edge_flip_comparison.rds", experiment_id="score_rules/edge_flip"))
}, 
"edge_flip_comparison", 
"Edge flip score values for common trajectory topologies", 
15, 
5
)
```

>>>>

This is the first overview about ~~deep reinforcement learning~~ trajectory inference publicly available online. It is comprehensive. Comments and criticisms are welcome.

@@

```{r}
add_sfig({
n_methods_over_time <- read_rds(figure_file("n_methods_over_time.rds", experiment_id="4-method_characterisation"))
trajectory_components_over_time <- read_rds(figure_file("trajectory_components_over_time.rds", experiment_id="4-method_characterisation"))

@@

<!-- history of Ti, different methods -->
```{r}
fixed_linear <-methods$name[which(methods$fixes_structure != 'free' & methods$maximal_trajectory_type == 'undirected_linear')]
fixed_bifurcating <-methods$name[which(methods$fixes_structure != 'free' & methods$maximal_trajectory_type == 'simple_fork')]
fixed_other <-methods$name[which(methods$fixes_structure != 'free' & methods$maximal_trajectory_type != 'undirected_linear'& methods$maximal_trajectory_type != 'simple_fork')]
fixed_algorithm <-methods$name[which(methods$fixes_structure == 'algorithm')]
fixed_parameters <-methods$name[which(methods$fixes_structure == 'parameter')]
```

Dozens of trajectory inference methods have been developed over the last years, and more are being published almost every month (`r ref("sfig", "methods_over_time", "a")`). Initially, most trajectory inference methods fixed the trajectory topology to a linear `r cite_methods(fixed_linear)`, bifurcating `r cite_methods(fixed_bifurcating)` or other `r cite_methods(fixed_other)` topology, by either algorithmic limitations `r cite_methods(fixed_algorithm)` or through user parameters `r cite_methods(fixed_parameters)`. The focus of these methods was therefore mainly to correctly ordering the cells along this fixed topology. However, several recent methods now infer both the ordering and the topology at the same time, although some topological limitations are still imposed (`r ref("sfig", "methods_over_time", "b")`). Methods which do not impose limitations on the topology at all are now being developed [@welch_slicer:_2016; @wolf_graph_2017], and it is expected that in the future methods will be able to model even more complex behavior, such as multiple dynamic processes happening at parallel in a single-cell or the integration of datasets from different patients [@tanay_scaling_2017; @cannoodt_computational_2016].

<!-- Common framework of TI, different components, naturally progressing towards the need for an evaluation -->
Although trajectory inference methods use a variety of algorithms and subcomponents to reach a final ordering, most consist of the following three steps: (i) preprocessing (normalization, filtering of genes and/or cells), (ii) conversion to a simplified representation using dimensionality reduction, clustering or graph building and (iii) ordering the cells along the simplified representation [@cannoodt_computational_2016]. Depending on the way the cells are ordered, some prior information about the dynamic process can be optionally used or required, which can both bias the trajectory to current knowledge but also guide the method towards choosing the right trajectory among many others. Furthermore, with the ongoing strong scalability in single-cell -omics technologies, it becomes more and more important that analysis methods become both scalable [@aibar_scenicsinglecellregulatory_2017] and user friendly.

@@

We gathered a list of `r sum(methods$is_ti, na.rm=TRUE)` TI methods, and selected a subset of `r sum(methods$evaluated)` for evaluation, primarily based on their free availability and the presence of a programming interface (`r ref("table", "methods")`). We characterized all methods in four different ways: possible trajectory topologies, implementation quality, prior information and the underlying algorithm (`r ref("fig", "4-method_characterisation")`).

`r ref("table", "methods", anchor=TRUE)` Overview of current trajectory inference methods, and whether they were included in this current evaluation study
```{r}
read_rds(figure_file("methods_table.rds", "4-method_characterisation"))
```

At the level of trajectory topologies, most recent methods can handle one or more splits (either bifurcations or multifurcations) in the trajectory (`r ref("fig", "4-method_characterisation", "a")`). However, only a handful, mostly recent, methods can currently handle more complex graph topologies which can include loops (`r ref("sfig", "methods_over_time", "b")`), one of which is included in the current evaluation (SLICER).

@@

```{r}
add_stable(
read_rds(figure_file("qc_checks.rds","4-method_characterisation")), 
"qc_checks", 
""
)
```

`r anchor("fig", "4-method_characterisation")`
![](`r figure_file("method_characteristics.svg", experiment_id="4-method_characterisation")`)
`r ref("fig", "4-method_characterisation", anchor=TRUE)` First characterization of each method.

```{r}
add_sfig({
read_rds(figure_file("method_qc_check.rds", experiment_id="4-method_characterisation"))
}, 
"method_qc_check", 
"Caption", 
10, 
13
)
```

Most methods fulfill most of the basic criteria, such as free availability and basic code quality criteria (`r ref("fig", "4-method_characterisation", "b")`). Recent methods tend to have higher code quality (`r ref("sfig", "qc_over_time")`). However, several aspects are consistently lacking for the majority of the methods (`r ref("fig", "4-method_characterisation", "e")`) and we believe that these should receive extra attention from developers. Although these outstanding issues cover all five categories, code assurance and documentation in particular are problematic areas (`r ref("sfig", "method_qc_check")` and `r ref("fig", "4-method_characterisation", "c")`), notwithstanding several studies pinpointing these as good practices [@wilson_best_2014; @artaza_top10metrics_2016].

```{r}
add_sfig({
read_rds(figure_file("qc_over_time.rds", experiment_id="4-method_characterisation"))
}, 
"qc_over_time", 
"Quality control scores and number of citations over time", 
10, 
10
)
```

@@

```{r}
add_sfig({
read_rds(figure_file("method_components.rds", experiment_id="4-method_characterisation"))
}, 
"method_components", 
"TI methods frequently share steps", 
15, 
7.5
)
```
## Evaluating the accuracy of the inferred trajectory

@@

```{r}
add_stable(
read_rds(figure_file("datasets.rds","2-dataset_characterisation")), 
"datasets", 
"Real datasets used in this study"
)
```

# Discussion

Trajectory inference is unique among most other categories of single-cell analysis methods, such as clustering, normalisation and differential expression, because it models the data in a way that was almost impossible using bulk data. Indeed, it is for this category that most methods were developed among other single-cell analysis categories, both according to omictools.org [@henry_omictoolsinformativedirectory_2014] (https://omictools.com/single-cell-rna-seq-category) and the "awesome single cell software" list [@davis_awesomesinglecelllistsoftware_2018] created by Sean Davis. In this preprint, we summarise the initial version of our evaluation of trajectory inference methods, focussing on the quality control

Limitations of the model:

```
* Robrecht.Cannoodt @robrechtc 10:21
en heterogene output van alle methode... true, maar sommigen hebben we wel wat moeten tweaken 😛
en ik kan me output bedenken dat niet in ons model zou kunnen passen
* Wouter Saelend @wouters 10:21
oh 😛
* Robrecht.Cannoodt @robrechtc 10:21
welja
* Wouter Saelend @wouters 10:21
Misschien een kritiekpunt is onzekerheid... maarja 😛
* Robrecht.Cannoodt @robrechtc 10:22
onzekerheid, en dat een cell in meerdere niet-neighbouring milestones zou kunnen zitten
als in, biologisch gezien houdt het mss geen steek, maar een methode zou wel kunnen zeggen, "wete, ik dénk dat het ofwel een DC is of een rode bloedcel"
* Wouter Saelend @wouters 10:24
hmmm uhu true, hoort allemaal wat samen met de onzekerheid. Je zou per cell een probabiliteitsdistributie doorheen het milestone netwerk kunnen geven, maar sla dat eens op 😛
* Robrecht.Cannoodt @robrechtc 10:25
uhu ^^
* Wouter Saelend @wouters 10:25
En bereken dan eens de emd 😛
* Robrecht.Cannoodt @robrechtc 10:25
idd...
deze discussie zou wel ergens een plaatsje kunnen krijgen in het manuscript, vind ik
* Wouter Saelend @wouters 10:25
Uhu in de discussie dan wss
* Robrecht.Cannoodt @robrechtc 10:26
tiens, een discussie die in de discussie zou plaats vinden, wat een goed idee 😮
<tease>
```

@@

```{r}
add_sfig({
read_rds(figure_file("trajectory_type_trees.rds", experiment_id="2-dataset_characterisation/2-trajectory_types"))
}, 
"trajectory_type_tree", 
"Different trajectory types can be classified in a directed acyclic graph, in which certain types are a generalisation of other trajectory types.", 
15, 
10
)
```

@@

* Methods returning a single pseudotime value: `r conversion("linear")`. A single edge between two milestones, its percentage proportional to the pseudotime.
* Methods which assign each cell to a branch together with a pseudotime value and a branch network: `r conversion("branching_local")`. Similarly, the branch network is used as the milestone network, the percentages of a cell are proportional with its branch pseudotime.
* Methods returning both a global pseudotime value and a probability for every end state: `r conversion("branching_global")`. We use a single start state and add an edge to every end milestone each representing an end state. The global pseudotime then determines the distance from the begin milestone, the rest of the cell's position is calculated by distributing the residual percentage over the end states, proportionally to the end state probabilities.
* Methods returning a cluster assignment and a cluster network:  `r conversion("branching_cluster")`. The cluster network was used as milestone network, each cell received percentage 1 or 0 based on its cluster assignment.
* Methods returning a cluster assignment, cluster network and dimensionality reduction: `r conversion("branching_local_projection")`. We projected each cell on the closest point on the edges between its own cluster and neighbouring clusters within the dimensionality reduction. This projection allowed us to give each cell a pseudotime within the edge, which was then converted into our model as described above, using the cluster network as milestone network.

@@

* DPT: We projected the cells onto the cluster network, consisting of a central milestone (this cluster contains the cells which were assigned to the "unknown" branch) and three terminal milestones, each corresponding to a tip point. This was then processed as described above.
* Sincell: To constrain the number of milestones this method creates, we merged two cell clusters iteratively until the percentage of leaf nodes was below a certain cutoff, default at 25%.  This was then processed as described above.
* SLICER: Although this method returns a branch network and a global pseudotime, it was not clear to us how to assign cells to individual branches in the network. We therefore used the cell graph provided by the method to calculate the shortest paths between each branch, and then projected other cells onto this backbone graph structure, creating a local pseudotime for every cell within its branch. The output was then further processed as above.
* SLICE: As discussed in the vignette of SLICE, we ran principal curves one by one for every edge detected by SLICE. This local pseudotime was then processed as above.
* MFA: We used the branch assignment as state probabilities, which together with the global pseudotime were processed as described above.

@@

```{r}
add_sfig({
read_rds(figure_file("modulenets.rds", experiment_id="2-dataset_characterisation/3-synthetic"))
}, 
"modulenets", 
"Module networks, milestone networks and the module dynamics for each of the different types of synthetic data", 
15, 
5
)
```

@@

```{r}
add_stable(
read_rds(figure_file("samplers.rds","2-dataset_characterisation/3-synthetic")), 
"samplers", 
"Distributions from which each parameter in the thermodynamic model was sampled"
)
```

@@

```{r}
add_sfig({
read_rds(figure_file("score_aspects.rds", experiment_id="3-score_rules/5-score_aspects"))
}, 
"score_aspects", 
"Different scores assess different aspects of the ", 
15, 
8
)
```

@@

```{r}
add_sfig({
read_rds(figure_file("edge_flip_comparison.rds", experiment_id="3-score_rules/4-edge_flip"))
}, 
"edge_flip_comparison", 
"Edge flip score values for common trajectory topologies", 
15, 
5
)
```
